---
title: "`interpret()` Usage Patterns"
author: 
 - name: Matthias Kloft
   orcid: 0000-0003-1845-6957
date: "`r Sys.Date()`"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    source: repo
    code-tools: true
    df-print: paged
execute:
  warning: true
  message: true
---

## Overview

This document demonstrates the different usage patterns for `interpret()`.
The `interpret()` function is the main function of the **psychinterpreter** package and can be used in multiple ways. It will automatically detect the type of input provided and run the appropriate method for interpretation.


## Setup

### Packages
```{r setup}
packages <- c(
  "psychinterpreter",
  "ellmer",
  "ollamar",
  "psych",
  "lavaan",
  "mirt",
  "dplyr",
  "here",
  "DT",
  "pander"
)
# Load required packages using pacman
if (!require("pacman")) install.packages("pacman")
pacman::p_load(char = packages)

```


## Run Factor Analysis
```{r run-fa}
fa_model <- fa(mtcars[,1:4], nfactors = 2, rotate = "oblimin")

var_info <- data.frame(
  variable = c("mpg", "cyl", "disp", "hp"),
  description = c("Miles per gallon", "Cylinders", "Displacement", "Horsepower")
)
```

## Usage Patterns

### Pattern 1: Fitted Model Object

Set provider and model:
```{r}
provider <- "anthropic"
llm <- "claude-sonnet-4-5"
```


Automatically extracts model components from fitted objects.
```{r pattern-1}
#| output: false
result <- interpret(
  fit_results = fa_model,
  variable_info = var_info,
  llm_provider = provider,
  llm_model = llm,
  word_limit = 50,
  silent = 1
)
```


### Pattern 2: Structured List

For custom data structures or manual extraction.

For FA, provide loadings (required) and optionally factor_cor_mat:

```{r pattern-2}
#| output: false
loadings_matrix <- fa_model$loadings
factor_cor_mat <- fa_model$Phi

result <- interpret(
  fit_results = list(loadings = loadings_matrix, factor_cor_mat = factor_cor_mat),
  variable_info = var_info,
  analysis_type = "fa",
  llm_provider = provider,
  llm_model = llm,
  word_limit = 50,
  silent = 1
)
```


### Pattern 3: Chat Session (Token-Efficient)

Reuse chat session across analyses to save tokens.

```{r pattern-3}
chat <- chat_session(analysis_type = "fa",
                     llm_provider = provider,
                     llm_model = llm)
result1 <- interpret(
  chat_session = chat,
  fit_results = fa_model,
  variable_info = var_info,
  word_limit = 50,
  silent = 1
)
result2 <- interpret(
  chat_session = chat,
  fit_results = list(loadings = loadings_matrix, factor_cor_mat = factor_cor_mat),
  variable_info = var_info,
  word_limit = 50,
  silent = 1
)
```

We can check the total global token usage:
```{r total-usage}
print(chat)
```


# Session Info
```{r session-info}
pander::pander(sessionInfo())
```

